Whose Truth is it Anyway? An Experiment on Annotation Bias in Times of Factual Opinion Polarization

### 2024-07-24 report/draft.Rmd
Updated figures, small changes intro and abtract

thanks: "Corresponding author: MACGvdV, Replication files are available on the author's Github account (https://github.com/MarikenvdVelden/bias-experiment); Author contributions: a) designed the study: MACGvdV, MR, AF, FL, WvA, & KW; b) conducted the study: MACGvdV, FL & MR; c) data cleaning & analysis: MACGvdV; d) writing of the paper: MACGvdV, MR, AF, FL, & KW"
anonymous: FALSE
abstract: "Information shapes citizens' political decision-making. This process is amply studied by social scientists, who have human annotation as a crucial instrument in their toolkit. To address concerns regarding the validity and reliability of annotation tasks, we establish strict standards. Due to the democratization of data and the advances in NLP more data can be analyzed or classified, making these standards are more important than ever: An algorithm trained on biased data will reproduce and often exacerbate bias. Our tools to create valid and reliable annotation data currently do not allow for dealing with different perspectives of annotators. In two pre-registered experiments in the United States and the Netherlands, we show that personal characteristics of annotators, like political ideology or knowledge, interfere with annotators' judgement of political stances. Our results show that to improve annotated data for automated text analyses, and for stance detection models in particular, we need to critically evaluate how we create our gold standards."
keywords: "Experiment, Annotation Bias, Ideology, Measurinig Political Position, Text-as-Data, Political Knowledge"
geometry: margin=1in
mainfont: cochineal
fontsize: 12pt
params:
  anonymous: ""
endnote: no
pandocparas: TRUE
sansitup: FALSE
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Crowd-coding platforms have been shown to be reliable platforms to collect a gold standard [@van2021validity].
Yet, from experimental studies [@clifford2015samples; @huff2015these; @berinsky2014separating] we know that these online platforms are populated by people that are "unlike" the general public, being younger and holding more liberal values.
This could mean that we introduce _sampling bias_ into the creation of the gold standard if we stricly consider only one position to be true.
@ennser2018impact report that coders of political texts incorporate prior beliefs about parties' issue stances into their coding decisions.
The authors find that party labels cue coders to a stance.
For example, coders are more likely to report a left-wing party to be pro-immigration and a populist right-wing party to be against based on the exact same sentence.
*Is this actually bias or a diversity of viewpoints?*
*And how big of a problem is this bias/diversity of viewpoints?*

In this context, bias would occur if there is one objectively correct solution to a coding task and coders systematically get the answer *wrong* due to prior beliefs. Opposed to this, diversity of viewspoints rather indicates that there are actually multiple ways of assessing the situation which all can be (in some way, depending on the siutation) correct. 
There is a long history of text annotation in studies analyzing political text.
Yet, the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks, research fields differ in their interpretations of what counts as a *correct* solution depending on the main goal of annotation.
Do we want to create a clean benchmark data set that follows strict linguistic rules?
Or do we want one that reflects how statements are being interpreted by "real" readers, how the perception of different types of respondents differs for the same stimulus? This could potentially affect downstream tasks as manual stance annotations are further used for e.g. sentiment analysis or frame analysis. Are we modelling the sentiment and frames of a particular section of the population while the results would change with a different sample? 
This question also taps into newer developments related to data annotations. 
Increasingly, Large Language Models such as ChatGPT are being used by researchers [@gilardi2023chatgpt] as well as annotators themselves [@veselovsky2023artificial] to "simulate" human respondents for annotation tasks.
These models draw among other things from the gold standard data sets that have been created by researchers, perpetuating the specific biases that are present in them and not allowing for any further disagreement or diversity of viewpoints.

Trends of polarization have shown that people do interpret information according to their ideological position [e.g., @rekker2022understanding; @lee2021more].
Does this mean that some are right and others are wrong? 
Or is there an ideological difference in the ground truth?
These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations (for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd; @ying2022topics).
However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2023don].
This leaves us between a rock and a hard place.
*Is there a potential bias in annotators that we should account for?*

In this paper, we build upon the NLP literature on disagreement -- or bias -- in annotation [e.g., see @shen2021sounds; @geva2019we; @sommerauer2020penguin; @plank2014learning] and so-called perspectivism [@cabitza2023toward; @havens2022beyond] -- i.e. the adoption of methods that integrate the opinions and perspectives of the human subjects involved in the knowledge representation step of the machine learning processes [@cabitza2023toward].
<!--This literature puts forward that disagreement can occur because of differences in ideological position or political knowledge [@shen2021sounds; @alkiek2022classification; @joseph2021mis].-->
We test the extent to which disagreement takes place and for what type of stances.
In addition, we provide guidance on how to deal with diversity in conceptions and political heterogeneity when creating gold standard data. <!-- that nowadays potentially occurs in our sample of annotators.-->
We use two high-powered pre-registered experiments (see [here](link) and [here](link)) in the Netherlands -- a low-level polarized country -- and in the U.S. -- a high-level polarized country.
We test possible biases stemming from ideology and political knowledge for correctly and overinterpreting stances.
Moreover, based on the results of @ennser2018impact, we test whether masking the political actor to mitigates these biases.^[The data and research compendium is published on the [[main author's github page]](X) -- ommitted for the review process.]
Another offered solution by Webb-Williams et al. (2023) is to collect information on political positions of the annotators to potentially re-weight responses based on population proportions to create composite measures, which is the focus of our H1 and H2. 
<!--effect of ideological distance between the annotator and the political actor in the text (H1), the effect of overinterpretation based on political knowledge or ideological engagement (H2), and an offered solution of masking the political actor to mitigate the effects of ideology and knowledge (H3).-->
In the experiments, we vary the level of specification with which a political actor takes a stance -- a declarative sentence versus a sentence where with some knowledge on politics, the stance might be inferred -- as well as whether the political actor is shown or masked with putting `[ACTOR]` instead of the political party.
We do this for four different political issues: *Environment*, *Immigration*, *Tax Policy*, and *EU* (for the Dutch case) or *Foreign Policy* (for the American case).
This country selection not only allow us to showcase the scope conditions of disagreement due to different levels of political heterogeneity, but also differentiates between languages.
English is not only the most dominant language for computational text analysis in social science [@baden2022three; @dolinsky2023threegaps], crowd-coders do not need to be first language speakers, given the dominance of English in our daily lives.
This is different for Dutch, it is a language spoken by a smaller community, typically first language speakers, yet still an often-enough researched case in computational text analysis in the social science [@baden2022three; @dolinsky2023threegaps].

>>>>

thanks: "Corresponding author: MACGvdV, Replication files are available on the author's Github account (https://github.com/MarikenvdVelden/bias-experiment); Author contributions: a) designed the study: MACGvdV, MR, AF, FL, WvA, & KW; b) conducted the study: MACGvdV, FL & MR; c) data cleaning & analysis: MACGvdV; d) writing of the paper: MACGvdV, MR, AF, FL, & KW"
anonymous: FALSE
abstract: "Information shapes citizens' political decision-making. This process is amply studied by social scientists, for whom human annotation is a crucial instrument in their toolkit. To address concerns regarding the validity and reliability of annotation tasks, benchmark data sets are created using rigourous annotation procedures. Due to the democratization of data and the advances in NLP more data can be analyzed or classified, making these benchmarks more important than ever: An algorithm trained on biased data will reproduce and often exacerbate bias. Current tools to create valid and reliable annotation data do not allow for dealing with different perspectives of annotators. In two pre-registered experiments in the United States and the Netherlands, we show that personal characteristics of annotators, like political ideology or knowledge, interfere with annotators' judgement of political stances. Our results show that to improve annotated data for automated text analyses, and for stance detection models in particular, we need to critically evaluate how we create our gold standards."
keywords: "Experiment, Annotation Bias, Ideology, Measurinig Political Position, Text-as-Data, Political Knowledge"
geometry: margin=1in
mainfont: cochineal
fontsize: 12pt
params:
  anonymous: ""
endnote: no
pandocparas: TRUE
sansitup: FALSE
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Crowd-coding platforms have been shown to be reliable platforms to collect a gold standard [@van2021validity].
Yet, from experimental studies [@clifford2015samples; @huff2015these; @berinsky2014separating] we know that these online platforms are populated by people that are "unlike" the general public, being younger and holding more liberal values.
This could mean that we introduce _sampling bias_ into the creation of the gold standard if we stricly consider only one position to be true.
@ennser2018impact report that coders of political texts incorporate prior beliefs about parties' issue stances into their coding decisions.
The authors find that party labels cue coders to a stance.
For example, coders are more likely to report a left-wing party to be pro-immigration and a populist right-wing party to be against based on the exact same sentence.
This brings us to the main question addressed in this paper:
*Is systematic disagreement actually bias or a diversity of viewpoints?*
*And how big of a problem is this bias/diversity of viewpoints?*

In this context, bias would occur if there is one objectively correct solution to a coding task and coders systematically get the answer *wrong* due to prior beliefs. 
Opposed to this, diversity of viewspoints rather indicates that there are actually multiple ways of assessing the situation which all can be (in some way, depending on the situation) correct. 
There is a long history of text annotation in studies analyzing political text.
Yet, the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks, research fields differ in their interpretations of what counts as a *correct* solution depending on the main goal of annotation.
Do we want to create a clean benchmark data set that stays as close to the manifest meaning of the text as possible?
Or do we want one that reflects how statements are being interpreted by "real" readers, how the perception of different types of respondents differs for the same stimulus? 
This could potentially affect downstream tasks as manual stance annotations are further used for e.g. sentiment analysis or frame analysis. 
<!-- Are we modelling the sentiment and frames of a particular section of the population while the results would change with a different sample?  -->
<!-- This question also taps into newer developments related to data annotations.  -->
<!-- Increasingly, Large Language Models such as ChatGPT are being used by researchers [@gilardi2023chatgpt] as well as annotators themselves [@veselovsky2023artificial] to "simulate" human respondents for annotation tasks. -->
<!-- These models draw among other things from the gold standard data sets that have been created by researchers, perpetuating the specific biases that are present in them and not allowing for any further disagreement or diversity of viewpoints. -->

Trends of polarization have shown that people do interpret information according to their ideological position [e.g., @rekker2022understanding; @lee2021more].
Does this mean that some are right and others are wrong? 
Or is there an ideological difference in the ground truth?
These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations (for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd; @ying2022topics).
However, work in computational linguistics has shown that increased annotation instructions do not always increase quality [@parmar2023don].
This leaves us between a rock and a hard place:
*Is there a potential bias in annotators that we should account for?*

In this paper, we build upon the NLP literature on disagreement -- or bias -- in annotation [e.g., see @shen2021sounds; @geva2019we; @sommerauer2020penguin; @plank2014learning] and so-called perspectivism [@cabitza2023toward; @havens2022beyond] -- i.e. the adoption of methods that integrate the opinions and perspectives of the human subjects involved in the knowledge representation step of the machine learning processes [@cabitza2023toward].
<!--This literature puts forward that disagreement can occur because of differences in ideological position or political knowledge [@shen2021sounds; @alkiek2022classification; @joseph2021mis].-->
We test the extent to which disagreement takes place and for what type of stances.
In addition, we provide guidance on how to deal with diversity in conceptions and political heterogeneity when creating gold standard data. <!-- that nowadays potentially occurs in our sample of annotators.-->
We use two high-powered pre-registered experiments (see [here](link) and [here](link)) in the Netherlands -- a low-level polarized country -- and in the U.S. -- a high-level polarized country.
We test possible biases stemming from ideology and political knowledge for correctly and overinterpreting stances.
Moreover, based on the results of @ennser2018impact, we test whether masking the political actor to mitigates these biases.^[The data and research compendium is published on the [[main author's github page]](X) -- ommitted for the review process.]
Another offered solution by Webb-Williams et al. (2023) is to collect information on political positions of the annotators to potentially re-weight responses based on population proportions to create composite measures, which is the focus of our H1 and H2. 
<!--effect of ideological distance between the annotator and the political actor in the text (H1), the effect of overinterpretation based on political knowledge or ideological engagement (H2), and an offered solution of masking the political actor to mitigate the effects of ideology and knowledge (H3).-->
In the experiments, we vary the level of specification with which a political actor takes a stance 
<!-- -- a declarative sentence versus a sentence where with some knowledge on politics, the stance might be inferred --  -->
as well as whether the political actor is shown or masked. 
<!-- with putting `[ACTOR]` instead of the political party. -->
<!-- We do this for four different political issues: *Environment*, *Immigration*, *Tax Policy*, and *EU* (for the Dutch case) or *Foreign Policy* (for the American case). -->
This country selection not only allow us to showcase the scope conditions of disagreement due to different levels of political heterogeneity, but also differentiates between languages.
English is not only the most dominant language for computational text analysis in social science [@baden2022three; @dolinsky2023threegaps], crowd-coders do not need to be first language speakers, given the dominance of English in our daily lives.
This is different for Dutch, it is a language spoken by a smaller community, typically first language speakers, yet still an often-enough researched case in computational text analysis in the social science [@baden2022three; @dolinsky2023threegaps].
